A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-23 latex/main.tex
Update concordance methods section

See
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/fig\_mutational\_spectra.ipynb}{fig\_mutational\_spectra.ipynb}
notebook for details.

\subsubsection*{Comparison with recombination detection methods}
[TODO Merge in this text that was in the Results]

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango X lineages associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb,
we summarized the detection results for these Pango X lineages (if available)
reported in the original studies \cite{Alfonsi2024,Li2024CovRecomb} (STAR Methods).
For each method, we determined the Pango X lineages
for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

[TODO explain why we restrict to the Type I events]

For RecombinHunt, we obtained two sets of detection results of the Pango X lineages
from the supplementary materials of Alfonso et al.\cite{Alfonsi2024}.
One set of results is based on mutational profiles calculated from a global GISAID dataset,
and the other set from a global NextStrain dataset.
The breakpoint intervals identified using RecombinHunt are defined
as the left and right recombination-informative sites.
We adjusted their coordinates by adding one to the right coordinate
so that the intervals are right-exclusive.

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the "independent" recombination events
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
which correspond to the Pango X lineages (if available).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
For each of these Pango X lineages,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.
We adjusted the breakpoint intervals by adding one to the right coordinate
so that they are right-exclusive.

The parents of a Pango X inferred using a method are deemed concordant
if they have Pango lineage assignments
that are at least as specific as those proposed by the community.
The breakpoint interval of a Pango X inferred using a method is deemed concordant
if it overlaps with that proposed by the community by at least one base.
These breakpoint intervals are defined as ranges of possible site positions
where a breakpoint is inferred to occur.
To make these breakpoint intervals right-exclusive, we adjusted them
by subtracting one from the left coordinate and adding two to the right coordinate.

>>>>

See t he
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/fig\_mutational\_spectra.ipynb}{fig\_mutational\_spectra.ipynb}
notebook for details.

\subsubsection*{Concordance of recombination events}
We compare concordance rates of sc2ts in identifying the 
Pango lineages of recombinant parents and the location
of breakpoint intervals with RecombinHunt\cite{Alfonsi2024} and 
CovRecomb\cite{Zhou2023virusrecom}.
We use these two methods because detection
results on the Pango X lineages were reported and made available 
in the original publications. We do not compare here with 
UShER+RIPPLES+RIVET because curated data on the X lineages is 
not available for this pipeline (see Document \ref{sec:detailed_ripples}
and Table~\ref{tab:ripples_x_lineages}, however).
We use the data from the Pango designation community 
(\url{https://github.com/cov-lineages/pango-designation/})
as the ``truth'' when comparing performance in identifying
parent Pango lineages and breakpoints between the methods.
We focus on the 16 Type I recombination events 
(Table~\ref{tab:pango_x_lineages}) identified by sc2ts for simplicity.
While this is certainly biased in favour of sc2ts (i.e., by 
ignoring likely sc2ts false negatives such as XB), it helps to 
understand the probable accuracy of sc2ts on the true positives.
It is important to emphasise that the ``true'' truth set among
the Pango X lineages is currently unknown, and there is significant 
disagreement among methods beyond a core set of lineages 
(see Discussion).

For RecombinHunt, we obtained two sets of detection results of the Pango X lineages
from the supplementary materials of Alfonso et al.\cite{Alfonsi2024}.
One set of results is based on mutational profiles calculated from a global GISAID dataset,
and the other set from a global NextStrain dataset.
The breakpoint intervals identified using RecombinHunt are defined
as the left and right recombination-informative sites.

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the ``independent'' recombination events
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
which correspond to the Pango X lineages (if available).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(\url{https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv}).
For each of these Pango X lineages,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.

The parents of a Pango X inferred using a method are deemed concordant
if they have Pango lineage assignments
that are at least as specific as those proposed by the Pango designation community.
The breakpoint interval of a Pango X inferred using a method is deemed concordant
if it overlaps with that proposed by the community by at least one base.

The results are reported in Table~\ref{tab:method_concordance}.
See the
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/tab\_methods\_concordance.ipynb}{tab\_methods\_concordance.ipynb}
notebook for details.
